DRAFT_Cleaning

### 2024-12-10 Draft_Cleaning.Rmd
split by rivers for PCA, clean PCA plots, got rid of unneeded doc

```{r setup, include=FALSE}
library(readr)
library(tidyverse)
library(lubridate)
library(caret)
library(ggfortify)
library(ggformula)
library(DiagrammeR)

dredge_data<-read_csv("DredgeData.csv")
survey_data<-read_csv("SurveyData.csv")
gage_data<-read_csv("UMR_IWW_1999_2021.csv")

gage_data2<- gage_data |>
  group_by(Date)|>
   summarize(across(everything(),~first(na.omit(.)), .names = "{.col}"))|>
  mutate(DATE_START= as_date(dmy(Date)))|>
    arrange(DATE_START)

gage_data2<-gage_data2|>
    mutate(across(c(-Date,-DATE_START),
                ~ ifelse(is.na(.),
                         coalesce((lag(.) + lead(.)) / 2, lag(.), lead(.)),
                         .)))|>
  select(-Date)

survey_Data<- survey_data |>
  filter(SURVEYTYPE == "conditional",
         SURVEY_YEAR >= 1999 & SURVEY_YEAR <= 2022,
          RIVER == "Mississippi_River" | RIVER == "Illinois_Waterway",
         SURVEYJOBIDPK == str_replace_all(SURVEYJOBIDPK, "IL", "IW"),
         !POOL %in% c("LP","CS"))|>
  mutate(VOLUMEDREDGED = 0,
         DREDGINGPURPOSE = "None",
         DREDGE_TYPE = "None",
         LABOR_TYPE = LABOR_SOURCE,
         EXECYEAR = SURVEY_YEAR,
         FEATURENAME = SDSFEATURENAME,
         DATE_START = as.character(DATE_START),
         DATE_START = as_date(DATE_START)
         )|>
  select(SURVEYJOBIDPK,EXECYEAR,DATE_START,DREDGINGPURPOSE,VOLUMEDREDGED, RIVER,POOL, DREDGE_TYPE,UP_RIV_MIL,DN_RIV_MIL)

dredge_data_cleaning <- dredge_data |>
   filter(
          DREDGINGPURPOSE != "harbor",
           EXECYEAR <= 2022)|>
  mutate(
    river_code = str_sub(PREHYDROSURVEY_ID, 1, 2),
    pool_code = str_sub(PREHYDROSURVEY_ID, 4, 5),
    district_code = str_sub(PREHYDROSURVEY_ID, 7, 9),
    river_mile = str_sub(PREHYDROSURVEY_ID, 23, 30),
    year_iteration = str_sub(PREHYDROSURVEY_ID, 32,34),
    SURVEYJOBIDPK = ifelse(
      str_detect(PREHYDROSURVEY_ID, "MVRUM|MVRIW"),
      PREHYDROSURVEY_ID,
      paste0(district_code, river_code, pool_code, "_", river_mile, "_", year_iteration)  
    ),
      SURVEYJOBIDPK = str_replace_all(SURVEYJOBIDPK, "IL", "IW"),
     DATE_START = as_date(mdy_hm(DATE_START)),
    SDSFEATURENAME = FEATURENAME)|>
  select(SURVEYJOBIDPK,EXECYEAR,DREDGINGPURPOSE, VOLUMEDREDGED, RIVER,POOL, DREDGE_TYPE,DATE_START, UP_RIV_MIL,DN_RIV_MIL)

combined_data <- bind_rows(survey_Data, dredge_data_cleaning)

@@

gage_survey <- combined_data |>
  left_join(gage_data2, by = "DATE_START")

@@

gage_survey_full <- gage_survey_filt|>  
  left_join(gage_data2, by = c("lead_7" = "DATE_START"), suffix = c("", "_7"))|>
  left_join(gage_data2, by = c("lead_14" = "DATE_START"), suffix = c("","_14"))

@@

### Create factored season variable that could be used in analysis
date_to_season <- function(date) {
  month <- as.numeric(format(date, "%m"))
  day <- as.numeric(format(date, "%d"))
  # Define the seasons based on date ranges
  if ((month == 3 && day >= 20) || (month >= 4 && month <= 6) || (month == 6 && day <= 20)) {
    return("Spring")
  } else if ((month == 6 && day >= 21) || (month >= 7 && month <= 9) || (month == 9 && day <= 21)) {
    return("Summer")
  } else if ((month == 9 && day >= 22) || (month >= 10 && month <= 12) || (month == 12 && day <= 20)) {
    return("Fall")
  } else {
    return("Winter")
  }
}

@@

@@

autoplot(dredge_PCA)

@@

ctrl = trainControl(method = "cv", number = 10) 
fit_dredge = train(VOLUMEDREDGED ~ .,
                 data = gage_xgb,
                 method = "xgbTree",
               tuneGrid = expand.grid(nrounds = c (15,25,50,100), 
                                              max_depth = 1:10,
                                        eta =c(0.3), gamma = 0,
                                        colsample_bytree = 1,
                                        min_child_weight = 1, subsample = 1),
                 verbosity = 0,
                 trControl = ctrl,
                 na.action = na.exclude)

@@

import_matrix = xgb.importance(model = fit_dredge$finalModel)
import_matrix
xgb.plot.importance(import_matrix)
```
###Cross Validation of Model
```{r}
set.seed(123)
n = nrow(gage_xgb)
n_outer_folds = 10

@@

  ctrl = trainControl(method = "cv", number = 10)
  fit_dredge_DCV = train(VOLUMEDREDGED ~ .,
                       data = gage_xgb[in_train, ],
                       method = "xgbTree",
                       tuneGrid = expand.grid(nrounds = c (15,25,50,100), 
                                              max_depth = 1:10,
                                        eta =c(0.3), gamma = 0,
                                        colsample_bytree = 1,
                                        min_child_weight = 1, subsample = 1),
                       verbosity = 0,
                       trControl = ctrl,
                       na.action = na.exclude)

@@

## Well that didnt work! Lets try modeling between the two river systems 
```{r}
combined_data <- bind_rows(survey_Data, dredge_data_cleaning)

@@

Miss_full <- Miss_River|>  
  left_join(gage_data2, by = c("lead_7" = "DATE_START"), suffix = c("", "_7"))|>
  left_join(gage_data2, by = c("lead_14" = "DATE_START"), suffix = c("","_14"))

IWW_full <- Illinois_River|>  
  left_join(gage_data2, by = c("lead_7" = "DATE_START"), suffix = c("", "_7"))|>
  left_join(gage_data2, by = c("lead_14" = "DATE_START"), suffix = c("","_14"))

@@

>>>>

```{r setup, include=FALSE}
library(readr)
library(tidyverse)
library(lubridate)
library(caret)
library(ggfortify)
library(ggformula)
library(DiagrammeR)
```

Read in the data
```{r}
dredge_data<-read_csv("DredgeData.csv", show_col_types = FALSE)
survey_data<-read_csv("SurveyData.csv", show_col_types = FALSE)
gage_data<-read_csv("UMR_IWW_1999_2021.csv", show_col_types = FALSE)

### The gage data has multiple records across different gages per day, we need to
### combine the rows by date so that there's only one record for date across all gages.
### Since we're going to be doing a lot of work with the date column, lets format it as 
### a date object and arrange by DATE_START and remove the old Date column
gage_data<- gage_data |>
  group_by(Date)|>
   summarize(across(everything(),~first(na.omit(.)), .names = "{.col}"))|>
  mutate(DATE_START= as_date(dmy(Date)))|>
    arrange(DATE_START)|>
  select(-Date)

### Some of the gage data has observation gaps where the sensors may not have been
### working, lets interpolate gage values where there's data before and after a gap
### with a maximum window of 2
gage_data<-gage_data|>
    mutate(across(c(-DATE_START),
                ~ ifelse(is.na(.),
                         coalesce((lag(.) + lead(.)) / 2, lag(.), lead(.)),
                         .)))
```

Now lets clean the survey data. I only want conditional survey records between
1999 to 2000, these are going to act as my background data. I want to combine
these records with my dredge data. I'm also going to filter by river, remove
pools that are not of interest, clean up the SURVEYJOBIDPK, and create some 
columns to support my join. Dates are also formatted as date objects and some
columns are renamed to make it easier to join my surveys to my dredge data.
Finally, select the columns we need for our analysis to join to our dredging 
data. 
```{r}
survey_data<- survey_data |>
  filter(SURVEYTYPE == "conditional",
         SURVEY_YEAR >= 1999 & SURVEY_YEAR <= 2022,
          RIVER == "Mississippi_River" | RIVER == "Illinois_Waterway",
         !POOL %in% c("LP","CS"))|>
  mutate(VOLUMEDREDGED = 0,
         DREDGINGPURPOSE = "None",
         DREDGE_TYPE = "None",
         LABOR_TYPE = LABOR_SOURCE,
         EXECYEAR = SURVEY_YEAR,
         FEATURENAME = SDSFEATURENAME,
         DATE_START = as.character(DATE_START),
         DATE_START = as_date(DATE_START)
         )|>
  select(EXECYEAR,DATE_START,DREDGINGPURPOSE,VOLUMEDREDGED, RIVER,POOL, DREDGE_TYPE,
         UP_RIV_MIL,DN_RIV_MIL)
```

Lets clean the dredge data, we only want records that are channel dredging events
and since the dredge records only go back to 1999, we just need to filter out data beyond 2022.
```{r}
dredge_data <- dredge_data |>
   filter(
          DREDGINGPURPOSE != "harbor",
           EXECYEAR <= 2022)|>
  mutate(
     DATE_START = as_date(mdy_hm(DATE_START)))|>
  select(EXECYEAR,DREDGINGPURPOSE, VOLUMEDREDGED, RIVER,POOL, DREDGE_TYPE,
         DATE_START, UP_RIV_MIL,DN_RIV_MIL)
```

Combine the dredge data and survey data into 1 table. The survey data will act as background/non-dredge data.
```{r}
combined_data <- bind_rows(survey_data, dredge_data)

@@

gage_survey <- combined_data |>
  left_join(gage_data, by = "DATE_START")

@@

gage_survey_full <- gage_survey_filt|>  
  left_join(gage_data, by = c("lead_7" = "DATE_START"), suffix = c("", "_7"))|>
  left_join(gage_data, by = c("lead_14" = "DATE_START"), suffix = c("","_14"))

@@

Create factored season variable that could be used in analysis
```{r}
date_to_season <- function(date) {
  month <- as.numeric(format(date, "%m"))
  day <- as.numeric(format(date, "%d"))
  # Define the seasons based on date ranges
  if ((month == 3 && day >= 20) || (month >= 4 && month <= 6) || (month == 6 && day <= 20)) {
    return("Spring")
  } else if ((month == 6 && day >= 21) || (month >= 7 && month <= 9) || (month == 9 && day <= 21)) {
    return("Summer")
  } else if ((month == 9 && day >= 22) || (month >= 10 && month <= 12) || (month == 12 && day <= 20)) {
    return("Fall")
  } else {
    return("Winter")
  }
}

@@

dredge <-gage_survey_cleaned |>
  filter(DREDGINGPURPOSE != "None")
  hist(dredge$VOLUMEDREDGED)

  median(dredge$VOLUMEDREDGED)

hist(gage_survey_cleaned$VOLUMEDREDGED)

@@

autoplot(dredge_PCA,data = gage_survey_cleaned, 
         color = 'SEASON',loadings = TRUE,
         loadings.colour = 'black',
         loadings.label = TRUE, loadings.label.size = 4, 
         loadings.label.colour = "blue")

@@

ctrl = trainControl(method = "cv", number = 5) 
fit_dredge = train(VOLUMEDREDGED ~ .,
                 data = gage_xgb,
                 method = "xgbTree",
               tuneGrid = expand.grid(nrounds = c (15,25,50,100), 
                                              max_depth = 1:10,
                                        eta =c(0.3), gamma = 0,
                                        colsample_bytree = 1,
                                        min_child_weight = 1, subsample = 1),
                 verbosity = 0,
                 trControl = ctrl,
                 na.action = na.exclude)

@@

import_matrix = xgb.importance(model = fit_dredge$finalModel)
import_matrix
xgb.plot.importance(import_matrix)
```
###Cross Validation of Model
```{r}
set.seed(123)
n = nrow(gage_xgb)
n_outer_folds = 5

@@

  ctrl = trainControl(method = "cv", number = 5)
  fit_dredge_DCV = train(VOLUMEDREDGED ~ .,
                       data = gage_xgb[in_train, ],
                       method = "xgbTree",
                       tuneGrid = expand.grid(nrounds = c (15,25,50,100), 
                                              max_depth = 1:10,
                                        eta =c(0.3), gamma = 0,
                                        colsample_bytree = 1,
                                        min_child_weight = 1, subsample = 1),
                       verbosity = 0,
                       trControl = ctrl,
                       na.action = na.exclude)

@@

## Well that didnt work! Lets try modeling between the two river systems

### Cleaning the data 
```{r}
combined_data <- bind_rows(survey_Data, dredge_data_cleaning)

@@

Miss_full <- Miss_River|>  
  left_join(gage_data, by = c("lead_7" = "DATE_START"), suffix = c("", "_7"))|>
  left_join(gage_data, by = c("lead_14" = "DATE_START"), suffix = c("","_14"))

IWW_full <- Illinois_River|>  
  left_join(gage_data, by = c("lead_7" = "DATE_START"), suffix = c("", "_7"))|>
  left_join(gage_data, by = c("lead_14" = "DATE_START"), suffix = c("","_14"))

@@

## PCA
```{r}
